Density-dependent selection and the limits of relative fitness

### 2018-03-27 manuscript.tex
l

Here we argue that the simplified models of density-regulated growth mentioned above give a narrow view of density-dependent selection that is misleading in some important ways. These ultimately derive from their failure to account for the presence of ``reproductive excesses'', that is, an excess of juveniles that are more fragile than their adult counterparts \citep{turner1968population}. Reproductive excesses allow selection to be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles. Although similar ideas featured prominently in early debates on the regulation of population density (e.g. \citealt{nicholson_1954}), few attempts have been made to explicitly model density-dependent selection in the presence of reproductive excesses. \cite{nei1971fertility} proposed a model with an explicit reproductive excess as part of a discussion of Haldane's ``cost of selection'' \citep{haldane_1957}, but used an unusual model of competition which was defined for at most two different types. In more complex models with detailed age structure (as opposed to juvenile vs adult), it is usually assumed that the density of a ``critical age-group'' controls the population's response to crowding, but otherwise the view of density dependence is surprisingly similar to that found in the models mentioned above, which have no age structure \citep[Chap. 4]{charlesworth_1994}. For example, selection proceeds in the direction of increasing equilibrium density in the critical age-group \citep[pp. 148]{charlesworth_1994}. As a result, the extent to which reproductive excesses justify the use of constant relative finesses is unclear beyond heuristic verbal arguments.

@@

We then outline the selection properties of our new variable-density lottery model (section ``Density-dependent selection in the variable-density lottery''). Next, we discuss
selection on the ability to contest territories, which behaves like a pure constant-$N$, constant-$s$ trait, and discuss how this fits with the classical notion of $K$-selection  (section ``The response of density to selection; $c$-selection versus $K$-selection''). We then consider selection on traits that regulate population density (section ``Density-regulating traits and the threat of strong selection''), and conclude by contrasting the classical density-dependent selection literature with our results (``Discussion'').

@@

Each iteration, adults produce propagules which disperse at random, independently of distance from their parents, and independently of each other. We assume that each adult from type $i$ produces $b_i$ propagules on average, so that the mean number of $i$ propagules dispersing to unoccupied territories is $m_i=b_in_iU/T$. The parameter $b_i$ can be thought of as a measure of ``colonization ability'', which combines fecundity and dispersal ability \citep{levins_71,tilman_94}. Random dispersal is then modeled using a Poisson distribution $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$ for the number $x_i$ of $i$ propagules dispersing to any particular unoccupied territory, where $l_i=m_i/U$ is the mean propagule density in unoccupied territories. The total propagule density will be denoted $L=\sum_i l_i$.

>>>>

Here we argue that the simplified models of density-regulated growth mentioned above are misleading in important ways. These ultimately derive from their failure to account for ``reproductive excesses'', that is, an excess of juveniles that are more fragile than their adult counterparts \citep{turner1968population}. Reproductive excesses allow selection to be concentrated at a juvenile ``bottleneck'', potentially keeping the density of adults constant even if selection is strong. The reproductive excess concept featured prominently in early debates on the regulation of population density (e.g. \citealt{nicholson_1954}), and is also familiar to evolutionists, particularly in relation to Haldane's ``cost of selection'' \citep{haldane_1957,turner1968population}. Reproductive excesses are also implicit in foundational evolutionary-genetic models like the Wright-Fisher, where each non-overlapping generation involves the production of an infinite number of zygotes, a constant number $N$ of which are sampled to form the next generation of adults. Likewise, in the Moran model, a juvenile is available to replace a dead adult every iteration no matter how rapidly adults are dying, and as a result $N$ remains constant.

Nevertheless, few studies of density-dependent selection have incorporated reproductive excesses. This requires density-regulated population growth in which the excess is finite and, in general, density-dependent. Such models are potentially much more complicated than those with no excess (e.g. logistic) or an infinite excess (e.g. Wright-Fisher). Nei's competitive selection model incorporated a finite reproductive excess to help clarify the ``cost of selection'' \citep{nei1971fertility,nagylaki_1992}, but used an unusual representation of competition based on pairwise interactions defined for at most two different types, and was also restricted to equal fertilities for each type. In more complex models with detailed age structure (as opposed to simply juveniles vs adults), it is usually assumed that the density of a ``critical age-group'' controls the population's response to crowding. The focus is then on the density in the critical age group,  and the role of density dependence is surprisingly similar to that found in models without any age structure \citep[Chap. 4]{charlesworth_1994}. For example, selection proceeds in the direction of increasing equilibrium density in the critical age-group, akin to classical $K$-selection \citep[pp. 148]{charlesworth_1994}. This gives little insight into the role of reproductive excesses in stabilizing adult densities.

@@

@@

Each iteration, adults produce propagules which disperse at random, independently of distance from their parents, and independently of each other. We assume that each adult from type $i$ produces $b_i$ propagules on average, so that the mean number of $i$ propagules dispersing to unoccupied territories is $m_i=b_in_iU/T$. The parameter $b_i$ can be thought of as a measure of ``colonization ability'', which combines fertility and dispersal ability \citep{levins_71,tilman_94}. Random dispersal is then modeled using a Poisson distribution $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$ for the number $x_i$ of $i$ propagules dispersing to any particular unoccupied territory, where $l_i=m_i/U$ is the mean propagule density in unoccupied territories. The total propagule density will be denoted $L=\sum_i l_i$.
